Machine learning high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-12-28 fame_appendex1.tex
acknowledgements

\section{Phase-space trajectories}\label{appendix:fame1_trajectory_generation}
We can see the RAMBO algorithm as a map from the unit hypercube in some high dimension into the $n$-particle phase-space. A flat distribution in the hypercube maps to a flat distribution in the multi-particle phase-space. 
To generate our phase-space trajectories we pick two points at random in the unit hypercube and map the line between them using the RAMBO mapping. As a result, each point on the resulting trajectory has equal probability density in phase-space. 
Any other phase-space generator that smoothly maps the unit hypercube to a multi-particle phase-space could replace RAMBO in this procedure and could lead to trajectories with very different characteristics. One could for example imagine a 
sophisticated algorithm that only maps points close to the boundary of the hypercube to soft or collinear configurations. Using such an algorithm would have trajectories that avoid configurations with many particles more collinear or soft than 
the end-points of the trajectories. We find that with RAMBO the trajectories tend not to avoid difficult phase-space configurations and are therefore a good test of the extrapolation properties of our method.

@@

\end{document}

>>>>

\section{Phase-space trajectories}\label{appendix:fame1_trajectory_generation}
We can see the {\RAMBO} algorithm as a map from the unit hypercube in some high dimension into the $n$-particle phase-space. A flat distribution in the hypercube maps to a flat distribution in the multi-particle phase-space. 
To generate our phase-space trajectories we pick two points at random in the unit hypercube and map the line between them using the {\RAMBO} mapping. As a result, each point on the resulting trajectory has equal probability density in phase-space. 
Any other phase-space generator that smoothly maps the unit hypercube to a multi-particle phase-space could replace {\RAMBO} in this procedure and could lead to trajectories with very different characteristics. One could for example imagine a 
sophisticated algorithm that only maps points close to the boundary of the hypercube to soft or collinear configurations. Using such an algorithm would have trajectories that avoid configurations with many particles more collinear or soft than 
the end-points of the trajectories. We find that with {\RAMBO} the trajectories tend not to avoid difficult phase-space configurations and are therefore a good test of the extrapolation properties of our method.

@@

Here we present another random phase-space trajectory similar to the one
in Figure~\ref{fig:random_trajectory} to demonstrate that features of that
figure are not unique.
\begin{figure}
    \includegraphics[width=\linewidth]{fame1/trajectory_1255.pdf}
    \caption{Another random phase-space trajectory where the `double+single'
    regions of phase-space are explicitly shown in blue.}
    \label{fig:trajectory_1255}
\end{figure}
\end{document}

### 2022-12-28 main.tex
update main and pdf
acknowledgements

\begin{document}
\title{Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders}
% \subtitle{An optional subtitle}
\author{Henry Truong}
\researchgroup{Institute for Particle Physics Phenomenology}
\maketitlepage*

\begin{abstract}
%
	This is some abstract about this thesis.
%
\end{abstract}

@@

	Research presented in this thesis is based on joint work
	published in the articles:
	\begin{itemize}
		\item Chapter \ref{chapter:fame1} is based on \cite{Maitre:2021uaa}: Daniel Ma\^{i}tre and Henry Truong, \textit{A factorisation-aware Matrix element emulator}, Journal of High Energy Physics, 2021 (11). 066.
		\item Chapter \ref{chapter:fame2} is based on ...
		\item Chapter \ref{chapter:fame3} is based on ...
	\end{itemize}

@@

\begin{acknowledgements*}
%
	I would like to thank\ldots
%
\end{acknowledgements*}

\begin{epigraph*}
%
	This a place for some profound quote which sums up your work, feelings or
	life.
	\source{Some book}{A. Author}
%
\end{epigraph*}

%\dedicationtext{This thesis is dedicated to}
\begin{dedication*}
%
	Someone special
	\also[and]{that other guy}
%
\end{dedication*}

>>>>

\begin{document}
\title{Machine learning high multiplicity matrix elements for electron-positron and hadron-hadron colliders}
% \subtitle{An optional subtitle}
\author{Henry Truong}
\researchgroup{Institute for Particle Physics Phenomenology}
\maketitlepage*

\begin{abstract}
%
	The LHC is a large-scale particle collider experiment
	collecting vast quantities of experimental data to study the fundamental
	particles, and forces, of nature. Theoretical predictions made with
	the SM can be compared with observables measured at experiments.
	These predictions rely on the use of Monte Carlo event generators to 
	simulate events which demand the evaluation of a matrix element.
	For high multiplicity processes this can take up a significant
	portion of the time spent simulating an event.
	In this thesis, we explore the usage of machine learning
	to accelerate the evaluation of matrix elements by introducing a
	factorisation-aware neural network model. Matrix elements are plagued with singular
	structures in regions of phase-space where particles become soft or collinear,
	however, the behaviour of the matrix element in these limits are well-understood.
	By exploiting the factorisation property of matrix elements in these limits,
	the model can learn how to best represent the approximation
	of the matrix elements as a linear combination of singular functions.
	We examine the application of the model to $e^{-}e^{+}$
	annihilation matrix elements at tree-level and one-loop level, as well
	as to leading order $pp$ collisions where the acceleration of event generation
	is critical for current experiments.
%
\end{abstract}

@@

	Research presented in this thesis is based on joint work:
	\begin{itemize}
		\item Chapter \ref{chapter:fame1} is based on \cite{Maitre:2021uaa}: Daniel Ma\^{i}tre and Henry Truong, \textit{A factorisation-aware Matrix element emulator}, Journal of High Energy Physics, 2021 (11). 066.
		\item Chapter \ref{chapter:fame2} is based on research with Daniel Ma\^{i}tre in preparation to be published.
		\item Chapter \ref{chapter:fame3} is based on research with Timo Janßen, Daniel Ma\^{i}tre, Steffen Schumann, and Frank Siegert in preparation to be published.
	\end{itemize}

@@

\begin{acknowledgements*}
%
	The work carried out during this journey would not have been
	possible without all the help and support from the people around me.
	Firstly, I would like to express my gratitude to Michael Spannowsky, and
	to my supervisor Daniel Ma\^{i}tre, who gave me a chance to undertake this
	doctorate when nobody else did.
	Daniel has always been supportive with his meticulous advice and constructive criticisms,
	and I thank him for giving me the freedom to explore different avenues during our joint research projects.

	All the people at the IPPP, especially Trudy and Joanne, have fostered a comfortable working
	and social environment of which I was very glad to be a part of.
	To all the students who have come and gone during my time at the IPPP, it was a pleasure
	to get to know you all. Special mentions go to Alan, Asli, Dorian, Francesco, Joe, Kevin, Marian,
	Ryan, and of course all the wonderful people in OC215 and beyond.
	Outside of the particles, shout-out to Aidan, Cameron, Ed, Jack, and Vicky for being great friends
	who I could always have a laugh with and chat about anything. I want to say thank you to Jack
	in particular for being a friend I could always count on for a good time, and a serious time.

	I want to express my deepest gratitude to my dear family and friends back home. Mum, dad, Annie,
	and Tracy, Chris and Tyrone, your endless love and support has been more helpful than you can imagine
	in completing this journey.

	These past four years of my life have been tremendously formative in many ways. All the people
	I have met, I have learned so much from, and I will cherish all
	the memories that we shared together. I will leave Durham as a better person.
%
\end{acknowledgements*}

### 2022-12-28 mc.tex
acknowledgements

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Monte Carlo Event Generators}\label{chapter:MC}
    At particle collider experiments,
    the collisions between incoming particles can lead
    to highly complex final-states with a large number
    of particles. In a hadron collider, such as the LHC,
    the particles we observe are not the constituents
    of the proton, the quarks and gluons. Instead we observe
    hadrons which are the bound states of these partons
    due to the property of confinement at low energy scales.

@@

    \section{Monte Carlo integration}\label{sec:MC_integration}
    As seen in Section~\ref{sec:xs}, the task of
    computing partonic cross-sections boils down
    to integrating the matrix elements over the
    final-state phase-space. Whilst it is possible to
    carry out these integrals analytically for the simplest
    cases, the integral quickly becomes intractable when the
    final-state multiplicity becomes large. For an $n$-body
    final-state the dimensionality of the integral scales as $d=3n-4$,
    where momentum conservation and on-shell conditions
    have already been applied to reduce the dimensionality.
    Therefore, instead of carrying out the integrals analytically,
    numerical methods become more attractive. The current
    prevailing method is Monte Carlo integration which
    has no scaling with the dimensionality of the integral,
    and so can be used for arbitrarily high multiplicity final-states
    with no performance penalty compared to lower multiplicities.
    The purpose of this chapter is to give a brief overview
    of Monte Carlo integration in the context of event
    generation, a more in-depth treatment of the
    specifics of Monte Carlo methods can be found in Ref.
    \cite{James:1980yn}.

    The basic idea of Monte Carlo integration is to approximate
    the integral of a function $f(\mathbf{x})$ by sampling it
    in $d$-dimensional parameter space points $\mathbf{x}$
    \begin{equation}\label{eqn:MC_integration}
        I = \int_{V} \mathrm{d}^{d}x \; f(\mathbf{x}) \approx \langle I(f) \rangle_{x} = V \dfrac{1}{N} \sum_{i=1}^{N} f(\mathbf{x}_{i}) = V \langle f \rangle_{x} \, ,
    \end{equation}
    where the points $\mathbf{x}$ are randomly,
    uniformly sampled in the volume $V$. $\langle f \rangle_{x}$
    denotes the average of $f$ over these uniformly sampled
    points. The law of large numbers ensures that the
    estimator $\langle I(f) \rangle_{x}$ will converge to
    the true value $I$ for $N \rightarrow \infty$. As a
    result of the central limit theorem, the function evaluations
    $f(\mathbf{x}_{i})$ will follow a Gaussian distribution with
    mean $\langle I(f) \rangle_{x}$,
    meaning an error estimate on the estimator is given by
    the standard error of the mean
    \begin{equation}\label{eqn:MC_error}
        \sigma_{I} = V \sqrt{\dfrac{\langle f^{2} \rangle_{x} - \langle f \rangle_{x}^{2}}{N}}
    \end{equation}
    which falls as $1/\sqrt{N}$, with no dependence
    on $d$.

    An algorithm that samples points $\mathbf{x}$ is called
    an integrator. To make use of the random numbers sampled,
    they need to be mapped to quantities that are useful in a simulation
    setting. The most commonly used mapping is to map
    the random numbers to four-momenta. An example of a
    simple integrator is the RAMBO algorithm \cite{Kleiss:1985gy,Platzer:2013esa}
    which samples four-momenta space uniformly. Since the integrator
    now effectively samples in momentum space, we refer to the points
    generated as phase-space points and so the terms integrator
    and phase-space generator are used interchangeably.

    Whilst Monte Carlo integration is independent of the
    dimensionality of the integral, meaning convergence is
    guaranteed for large enough $N$, there are variance
    reduction methods to more rapidly reduce the error
    on the estimator.

@@

    \section{Event unweighting}\label{sec:unweighting}
    Once a suitable configuration of momenta
    has been found by the integrator, this phase-space point forms the basis of an
    event. The matrix element of interest can be evaluated at
    this phase-space point at the desired order in perturbation
    theory (see Section~\ref{sec:me_generators}).
    Additionally, there will be an associated Jacobian
    from the mapping from random variables to the four-momenta.
    This Jacobian, along with the PDF weights and flux factors,
    are referred to as the phase-space weight, $J_{i}$. Taking
    this phase-space weight together
    with the matrix element forms the event weight\footnote{For
    the more sophisticated sampling methods employed in event generators
    there will be additional weights involved.}
    \begin{equation}\label{eqn:event_weight}
        w_{i} = |\mathcal{M}|_{i}^{2} J_{i} \, .
    \end{equation}
    The event weight can be interpreted as the probability
    of the event occurring at an experiment \cite{Plehn:2009nd}.
    By generating a number of events in this manner,
    and histogramming them, a sample of weighted events is created.

    At experiments, collision events are either observed or unobserved,
    meaning each observed configuration contributes equally to
    the measured cross-sections, and so are equally weighted.
    Therefore, in order to generate events that
    compare to experiments, it is necessary to translate
    weighted events into unit-weight events.
    This process is known as unweighting and is typically
    carried out using a rejection sampling algorithm, as
    outlined in Algorithm~\ref{alg:unweighting}. The rejection
    sampling algorithm makes use of $w_{\mathrm{max}}$ which is the
    maximal event weight in the integration volume.

    \begin{algorithm}
        \caption{Rejection sampling for unweighting events}\label{alg:unweighting}
        \While{unweighting}{
            generate random phase-space point $\mathbf{x}$\;
            evaluate event weight $w \gets w(\mathbf{x})$\;
            generate uniform random number $r \gets$ Random(0, 1)\;
            \If{$w / w_{\mathrm{max}} > r$}{
                \KwRet $\mathbf{x}$ and $\tilde{w} \gets 1$\;
                % $\tilde{w} \gets \max(1,\; w / w_{\mathrm{max}})$\;
            }
        }
    \end{algorithm}

    From this description, it is clear that the number
    of accepted events will generally be lower than
    the number of total events generated. This is encapsulated
    in the unweighting efficiency
    \begin{equation}\label{eqn:unweighting_efficiency}
        \varepsilon = \dfrac{N_{\mathrm{accepted}}}{N_{\mathrm{total}}} \, ,
    \end{equation}
    whose inverse is the average number of trial
    events required to obtain a single unweighted event.
    Unweighting efficiencies for high-multiplicity processes
    are often well below 1\% \cite{Hoche:2019flt,Gao:2020zvv} and the development
    of methods to improve the efficiency are an active area of research
    \cite{Jadach:1999sf,Jadach:2002kn}.
    These low efficiencies lead to most of the
    yearly event generator CPU budgets of LHC experiments to be spent on
    unweighting events \cite{HSFPhysicsEventGeneratorWG:2020gxw}.

    An alternative to improving the unweighting efficiency
    itself is to accelerate the process of unweighting itself. This
    was explored in Ref. \cite{Danziger:2021eeg} wherein a
    novel two-stage unweighting procedure employing a surrogate
    model of the exact event weight was used to decrease the amount of
    time spent unweighting events. This idea will be elaborated on
    in Chapter~\ref{chapter:fame3} where we build upon
    Ref. \cite{Danziger:2021eeg} by increasing the fidelity of
    the surrogate model.

@@

    As seen in Eq. (\ref{eqn:matrix_element}),
    the exact matrix element is approximated by a perturbative
    series in a coupling parameter. By now it is standard to
    have at least the LO and NLO contributions in this expansion.
    The calculations of
    tree-level \cite{Mangano:2002ea,Krauss:2001iv,Gleisberg:2008fv,Cafarella:2007pc,Alwall:2011uj,Kilian:2007gr}
    and one-loop \cite{Berger:2008sj,Cullen:2011ac,Cullen:2014yla,Bevilacqua:2011xh,Hirschi:2011pa,Alwall:2014hca,Frederix:2018nkq,Campbell:1999ah,Campbell:2021vlt,Badger:2012pg,Cascioli:2011va,Buccioni:2019sur,Actis:2016mpe,Denner:2017wsf}
    matrix elements are now largely fully automated
    in NLO QCD and electroweak corrections.
    Many of the modern matrix element generators are supplied
    with a common generic interface to event generators,
    the Binoth Les Houches Accord \cite{Binoth:2010xt,Alioli:2013nda}.
    For more information on the machinery developed to
    compute matrix elements at tree-level and
    one-loop level required for SM predictions, see Ref. \cite{Campbell:2017hsr}.
    NNLO and N$^{3}$LO QCD corrections have been computed
    for select processes but have yet to be reach the
    level of automation of NLO.
    For a review of recent advancements see Refs. \cite{Caola:2022ayt,Campbell:2022qmc}.

@@

    \subsection*{Parton shower}\label{sec:parton_showers}
    Fixed-order calculations of matrix elements are limited
    to relatively few final-state particles due to the complexity
    of analytical expressions and their phase-space integrals.
    They are also only accurate at high energy scales.
    To evolve from the fixed-order
    calculation to the non-perturbative regime where hadrons
    are formed, parton showers are employed in event generators.

@@

    \subsection*{Hadronisation}\label{sec:hadronisation}
    The simulation of the formation of hadrons in the low-energy regime
    of an event generator follows phenomenological, non-perturbative models.
    These hadronisation models require the combination
    of quarks and antiquarks to form colourless hadrons.
    The two main classes of models used to simulate
    hadronisation are the cluster models \cite{Webber:1983if,Winter:2003tt}
    and string models \cite{Andersson:1983ia}.
    For more discussion on hadronisation
    see Ref. \cite{Webber:1999ui}.

    \section{A note on theoretical uncertainties}\label{sec:scale_variations}
    The partonic cross-section, $\hat{\sigma}(\mu_{F}, \mu_{R})$, has
    a dependence on both the factorisation scale and the
    renormalisation scale. Although these scales are unphysical,
    and so any observable should not depend on them,
    there is a residual dependence on them in fixed-order
    calculations due to missing higher order terms.

    The most widely adopted approach to estimate the
    theoretical uncertainties associated with these
    missing higher order terms is to carry out a seven point
    scale variation. Typically in a calculation,
    $\mu_{0} = \mu_{F} = \mu_{R}$ where $\mu_{0}$ depends on the specifics
    of the process of interest. Taking the value of the
    partonic cross-section at this scale as the central value,
    the seven variations correspond to multiplying ($\mu_{F}, \, \mu_{R}$) by the factors 
    \begin{equation}\label{eqn:scale_variations}
        S \in \left\{(1, 1), (\sfrac{1}{2}, \sfrac{1}{2}), (2, 2), (\sfrac{1}{2}, 1), (1, \sfrac{1}{2}), (2, 1), (1, 2) \right\} \, ,
    \end{equation}
    and re-evaluating the cross-section at these scales.
    The deviations from the central cross-section value
    forms an estimate of the uncertainty in the perturbatively
    computed cross-section, when compared to the true cross-section.
    This can be written as
    \begin{equation}
        \hat{\sigma}(\mu_{F}, \mu_{R}) = \hat{\sigma}(\mu_{0}, \mu_{0}) \pm \delta \, ,
    \end{equation}
    where $\delta$ is determined from the scale variations.
    However, there is no single, well agreed upon definition
    of $\delta$ \cite{Cacciari:2011ze}.
    The rationale behind taking this approach is that $\delta$
    is approximately the same order as the true missing
    higher order terms.

>>>>

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Monte Carlo Event Generators}\label{chapter:MC}
    At particle collider experiments,
    the collisions between incoming particles can lead
    to highly complex final-states with a large number
    of particles. In any collider involving QCD processes,
    the particles we observe are not the constituents
    of the proton, the quarks and gluons. Instead we observe
    hadrons which are the bound states of these partons
    due to the property of confinement at low energy scales.

@@

    \section{Monte Carlo integration}\label{sec:MC_integration}
    As seen in Section~\ref{sec:xs}, the task of
    computing partonic cross-sections boils down
    to integrating the matrix elements over the
    final-state phase-space. Whilst it is possible to
    carry out these integrals analytically for the simplest
    cases, the integral quickly becomes intractable when the
    final-state multiplicity becomes large. For an $n$-body
    final-state the dimensionality of the integral scales as $d=3n-4$,
    where momentum conservation and on-shell conditions
    have already been applied to reduce the dimensionality.
    Therefore, instead of carrying out the integrals analytically,
    numerical methods become more attractive. The current
    prevailing method is Monte Carlo integration as the
    integration error has no scaling with the dimensionality of the integral,
    and so can be used for arbitrarily high multiplicity final-states
    with no performance penalty compared to lower multiplicities.
    The purpose of this chapter is to give a brief overview
    of Monte Carlo integration in the context of event
    generation, a more in-depth treatment of the
    specifics of Monte Carlo methods can be found in Ref.
    \cite{James:1980yn}.

    The basic idea of Monte Carlo integration is to approximate
    the integral of a function $f(\mathbf{x})$ by sampling it
    in $d$-dimensional parameter space points $\mathbf{x}$
    \begin{equation}\label{eqn:MC_integration}
        I = \int_{V} \mathrm{d}\mathbf{x} \; f(\mathbf{x}) \approx I_{N} = V \dfrac{1}{N} \sum_{i=1}^{N} f(\mathbf{x}_{i}) = V \langle f \rangle_{\mathbf{x}} \, ,
    \end{equation}
    where the points $\mathbf{x}$ are randomly,
    uniformly sampled in the volume $V$. Given a sample of $N$
    points $\{\mathbf{x}_{1}, \ldots, \mathbf{x}_{N}\} \in V$,
    $\langle f \rangle_{\mathbf{x}}$
    denotes the average of $f$ over these uniformly sampled
    points. The law of large numbers ensures that the
    estimator $I_{N}$ will converge to
    the true value $I$ for $N \rightarrow \infty$. As a
    result of the central limit theorem, the function evaluations
    $f(\mathbf{x}_{i})$ will follow a Gaussian distribution with
    mean $I_{N}$,
    meaning an error estimate on the estimator is given by
    the standard error of the mean
    \begin{equation}\label{eqn:MC_error}
        \sigma_{I} = V \sqrt{\dfrac{\langle f^{2} \rangle_{\mathbf{x}} - \langle f \rangle_{\mathbf{x}}^{2}}{N-1}}
    \end{equation}
    which falls as $1/\sqrt{N}$, with no dependence
    on $d$.

    An algorithm that samples points $\mathbf{x}$ is called
    an integrator. To make use of the random numbers sampled,
    they need to be mapped to quantities that are useful in a simulation
    setting. The most commonly used mapping is to map
    the random numbers to four-momenta. An example of a
    simple integrator is the {\RAMBO} algorithm \cite{Kleiss:1985gy,Platzer:2013esa}
    which samples four-momenta space uniformly in a phase-space
    integration. Since the integrator
    now effectively samples in momentum space, we refer to the points
    generated as phase-space points and so the terms integrator
    and phase-space generator are used interchangeably.

    Whilst Monte Carlo integration is independent of the
    dimensionality of the integral, meaning convergence is
    guaranteed for large enough $N$, there are variance
    reduction methods to more rapidly reduce the error
    on the estimator. These methods aim to reduce the numerator
    in Eq.~(\ref{eqn:MC_error}).

@@

    \section{Event unweighting}\label{sec:unweighting}
    Once a suitable configuration of momenta
    has been found by the integrator, this phase-space point forms the basis of an
    event. The matrix element of interest can be evaluated at
    this phase-space point at the desired order in perturbation
    theory (see Section~\ref{sec:me_generators}).
    Additionally, there will be an associated Jacobian
    from the mapping from random variables to the four-momenta.
    This Jacobian, along with the PDF weights and flux factors,
    are referred to as the phase-space weight, $J_{\mathrm{PS}}$. Taking
    this phase-space weight together
    with the matrix element forms the event weight\footnote{For
    the more sophisticated sampling methods employed in event generators
    there will be additional weights involved.}
    \begin{equation}\label{eqn:event_weight}
        w = |\mathcal{M}|^{2} J_{\mathrm{PS}} \, .
    \end{equation}
    The event weight can be interpreted as the probability
    of the event occurring at an experiment \cite{Plehn:2009nd}.
    By generating a number of events in this manner
    a sample of weighted events is created. Cross-sections
    and distributions can then be calculated by histogramming
    these weights.

    The generation of weighted events is generally undesirable
    as many events will have small weights with only a few events
    making sizeable contributions to predictions. Furthermore,
    these small weights still have to be run through expensive detector
    simulations, making the usage of weighted events highly inefficient.
    Therefore, it has become customary to generate events
    with equal relative weight, so-called unweighted events.

    An additional reason to generate unweighted events is that
    experimentalists would like the probability of events occurring
    at colliders to be described only by their frequency and not
    with any additional weights. Since unweighted events have
    equal relative weight, they satisfy this criteria.

    The process of generating unweighted events is known as
    unweighting and is typically carried out using a rejection
    sampling algorithm.
    Events are accepted or rejected by comparing the ratio
    $w / w_{\mathrm{max}}$ to a uniform random number, $r$,
    sampled in the interval $[0, 1]$, where $w_{\mathrm{max}}$ is
    the maximal event weight in the integration volume.
    Events are then accepted if $w / w_{\mathrm{max}} > r$.
    This can be interpreted as converting the event weights
    into a probability of being accepted or rejected as
    small event weights will have a low probability of being
    accepted, whereas the opposite is true for large event weights.
    The rejection algorithm is outlined explicitly in
    Algorithm~\ref{alg:unweighting}.

    \begin{algorithm}[ht]
        \caption{Rejection sampling for unweighting events}\label{alg:unweighting}
        \While{unweighting}{
            generate random phase-space point $\mathbf{x}$\;
            evaluate event weight $w \gets w(\mathbf{x})$\;
            generate uniform random number $r \gets$ Random(0, 1)\;
            \If{$w / w_{\mathrm{max}} > r$}{
                \KwRet $\mathbf{x}$ and $\tilde{w} \gets 1$\;
                % $\tilde{w} \gets \max(1,\; w / w_{\mathrm{max}})$\;
            }
        }
    \end{algorithm}

    From this description, it is clear that the number
    of accepted events will generally be lower than
    the number of total events generated. This is encapsulated
    in the unweighting efficiency
    \begin{equation}\label{eqn:unweighting_efficiency}
        \epsilon = \dfrac{N_{\mathrm{accepted}}}{N_{\mathrm{total}}} \, ,
    \end{equation}
    whose inverse is the average number of trial
    events required to obtain a single unweighted event.
    Unweighting efficiencies for high-multiplicity processes
    are often well below 1\% \cite{Hoche:2019flt,Gao:2020zvv} and the development
    of methods to improve the efficiency are an active area of research
    \cite{Jadach:1999sf,Jadach:2002kn}.
    These low efficiencies lead to most of the
    yearly event generator CPU budgets of LHC experiments to be spent on
    unweighting events \cite{HSFPhysicsEventGeneratorWG:2020gxw}.

    An alternative to improving the unweighting efficiency
    itself is to accelerate the process of unweighting itself. This
    was explored in Ref. \cite{Danziger:2021eeg} wherein a
    novel two-stage unweighting procedure employing a surrogate\footnote{The authors of Ref.~\cite{Danziger:2021eeg}
    refer to their model approximating the event weight as a surrogate model, whereas in
    this thesis we have used the term emulator. We will use these two terms interchangeably
    henceforth.}
    model of the exact event weight was used to decrease the amount of
    time spent unweighting events. This idea will be elaborated on
    in Chapter~\ref{chapter:fame3} where we build upon
    Ref. \cite{Danziger:2021eeg} by increasing the fidelity of
    the surrogate model.

@@

    As seen in Eq. (\ref{eqn:matrix_element}),
    the exact matrix element is approximated by a perturbative
    series in a coupling parameter. By now it is standard to
    have at least the LO and NLO contributions in this expansion.
    The calculations of
    tree-level \cite{Mangano:2002ea,Krauss:2001iv,Gleisberg:2008fv,Cafarella:2007pc,Alwall:2011uj,Kilian:2007gr}
    and one-loop \cite{Berger:2008sj,Cullen:2011ac,Cullen:2014yla,Bevilacqua:2011xh,Hirschi:2011pa,Alwall:2014hca,Frederix:2018nkq,Campbell:1999ah,Campbell:2021vlt,Badger:2012pg,Cascioli:2011va,Buccioni:2019sur,Actis:2016mpe,Denner:2017wsf}
    matrix elements are now largely fully automated
    in NLO QCD and electroweak corrections.
    Many of the modern matrix element providers are supplied
    with a common generic interface to event generators,
    the Binoth Les Houches Accord \cite{Binoth:2010xt,Alioli:2013nda}.
    For more information on the machinery developed to
    compute matrix elements at tree-level and
    one-loop level required for SM predictions, see Ref. \cite{Campbell:2017hsr}.
    NNLO and N$^{3}$LO QCD corrections have been computed
    for select processes but have yet to be reach the
    level of automation of NLO.
    For a review of recent advancements see Refs. \cite{Caola:2022ayt,Campbell:2022qmc}.

@@

    \subsection*{Parton shower}\label{sec:parton_showers}
    Fixed-order calculations of matrix elements are limited
    to relatively few final-state particles due to the complexity
    of analytical expressions and their phase-space integrals.
    They are also only accurate at high energy scales. For low
    transverse momentum emissions, large logarithms appearing at
    every order breaks the $\alpha_{s}$ perturbative expansion.
    To describe these kinematic regions, and to evolve from the fixed-order
    calculation to the non-perturbative regime where hadrons
    are formed, parton showers are employed in event generators.

@@

    \subsection*{Hadronisation}\label{sec:hadronisation}
    The simulation of the formation of hadrons in the low-energy regime
    of an event generator follows phenomenological models,
    which are non-perturbative and consist of many
    parameters fitted to data.
    These hadronisation models require the combination
    of quarks and antiquarks to form colourless hadrons.
    The two main classes of models used to simulate
    hadronisation are the cluster models \cite{Webber:1983if,Winter:2003tt}
    and string models \cite{Andersson:1983ia}.
    For more discussion on hadronisation
    see Ref. \cite{Webber:1999ui}.

    \section{Estimation of theoretical uncertainties}\label{sec:scale_variations}
    The partonic cross-section, $\hat{\sigma}(\mu_{F}, \mu_{R})$, has
    a dependence on both the factorisation scale and the
    renormalisation scale. Although these scales are unphysical,
    and so any observable should not depend on them,
    there is a residual dependence on them in fixed-order
    calculations due to missing higher order terms.

    The most widely adopted approach to estimate the
    theoretical uncertainties associated with these
    missing higher order terms is to carry out a seven point
    scale variation. Typically in a calculation,
    $\mu_{0} = \mu_{F} = \mu_{R}$ where $\mu_{0}$ depends on the specifics
    of the process of interest. Taking the value of the
    partonic cross-section at this scale as the central value,
    the seven variations correspond to multiplying ($\mu_{F}, \, \mu_{R}$) by the factors 
    \begin{equation}\label{eqn:scale_variations}
        S \in \left\{(1, 1), (\sfrac{1}{2}, \sfrac{1}{2}), (2, 2), (\sfrac{1}{2}, 1), (1, \sfrac{1}{2}), (2, 1), (1, 2) \right\} \, ,
    \end{equation}
    and re-evaluating the cross-section at these scales.
    The deviations from the central cross-section value
    forms an estimate of the uncertainty in the perturbatively
    computed cross-section, when compared to the true cross-section.
    This can be written as
    \begin{equation}
        \hat{\sigma}(\mu_{F}, \mu_{R}) = \hat{\sigma}(\mu_{0}, \mu_{0}) \pm \delta \, ,
    \end{equation}
    where $\delta$ is determined from the scale variations.
    However, there is no single, well agreed upon definition
    of $\delta$ \cite{Cacciari:2011ze}.
    The rationale behind taking this approach is that $\delta$
    is approximately the same order of magnitude as the true missing
    higher order terms.

### 2022-12-28 ml.tex
acknowledgements

\begin{document}
\chapter{Modern machine learning algorithms and their uses}
\label{chapter:ml}
\section{Rise of machine learning in HEP}
    In high energy physics, where analyses
    depend on making use of large computing clusters
    and international grid efforts \cite{WLCG}, advancements
    in technology and more efficient methods are
    required to keep up with the needs of the
    experimental and theoretical communities.
    However, the computational resources needed
    are at risk of outpacing the growth in these research
    areas \cite{Bothmann:2022thx}, especially with
    HL-LHC projected to begin operating at the end
    of this decade \cite{ZurbanoFernandez:2020cco}.
    Therefore new techniques and
    algorithms from machine learning 
    have been adopted to augment the ongoing efforts
    in high energy physics. Machine
    learning is a category of artificial intelligence
    concerned with the design of algorithms that
    automates learning from data.
    By now, machine learning is ubiquitous in
    particle physics analysis, from collection
    of data from experiments, all the way to novel
    applications on the theory side.

>>>>

\begin{document}
\chapter{Machine learning in high energy physics}
\label{chapter:ml}
\section{Rise of machine learning in high energy physics}
    In high energy physics, where analyses
    depend on making use of large computing clusters
    and international grid efforts \cite{WLCG}, advancements
    in technology and more efficient methods are
    required to keep up with the needs of the
    experimental and theoretical communities.
    However, the computational resources needed
    are at risk of outpacing the growth in these research
    areas \cite{Bothmann:2022thx}, especially with
    HL-LHC projected to begin operating at the end
    of this decade \cite{ZurbanoFernandez:2020cco}.
    Therefore new techniques and
    algorithms from machine learning 
    have been adopted to augment the ongoing efforts
    in high energy physics. Machine
    learning is a category of artificial intelligence
    concerned with the design of algorithms that
    automates learning from data.
    By now, machine learning is ubiquitous in
    particle physics analysis, from collection
    of data from experiments, all the way to novel
    applications on the theory side.
